\documentclass[10pt]{article}

\usepackage[T1]{fontenc}
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\usepackage{hyperref}
\usepackage{longtable}
\usepackage{titlesec}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{booktabs}
\usepackage{enumitem}
\usepackage{cite}
\setlist{noitemsep, topsep=3pt, partopsep=0pt}

\setstretch{1.15}

\titleformat{\section}{\normalfont\Large\bfseries}{\thesection}{1em}{}
\titleformat{\subsection}{\normalfont\large\bfseries}{\thesubsection}{1em}{}
\titleformat{\subsubsection}{\normalfont\normalsize\bfseries}{\thesubsubsection}{1em}{}

\lstset{
  basicstyle=\ttfamily\small,
  breaklines=true,
  frame=single,
  backgroundcolor=\color{gray!8},
  rulecolor=\color{gray!40},
}

\title{KGRAG: A Knowledge Compiler and Federated Retrieval Layer\\
for Ontologically Grounded Domains}

\author{Eric G. Suchanek, PhD\\
Flux-Frontiers, Liberty Township OH\\[4pt]
\small\url{https://github.com/Flux-Frontiers/KGRAG}}
\date{}

\begin{document}

\maketitle

\begin{abstract}
KGRAG is a federation and orchestration layer for structural knowledge graphs
derived from heterogeneous source domains: Python source code (CodeKG), natural
language documentation (DocKG), scientific or domain-specific corpora (MetaKG),
and a growing set of formally structured domains including biochemical pathway
data, protein structure files, legal codes, personal diary corpora, scriptural
texts, and biographical knowledge graphs. Each backend constructs its knowledge
graph from the formal structure of its source---abstract syntax trees for code,
document section and reference graphs for text, domain ontologies for scientific
data---without any language model participation in graph construction.

The central thesis of this work is that the power of structural knowledge-graph
retrieval scales directly with the formal completeness of the source domain's
ontology. Programming languages represent the ideal case: their grammars are
complete formal specifications, and the derived call graph is correct by
theorem. The same principle applies to biochemical pathway schemas, statutory
hierarchies in books of law, protein data bank file formats, and the structured
hierarchies of scripture and diary. KGRAG treats all of these domains
identically through a uniform five-method adapter protocol.

The design treats the structurally derived graph as ground truth and uses
semantic embeddings strictly as an acceleration layer for locating entry points
into that graph. Graph traversal, ranking, and snippet extraction are
deterministic: identical inputs produce identical outputs, and no component of
the structural pipeline produces probabilistic results. When KGRAG output is
passed to a large language model for synthesis, the model receives verified
facts with full source provenance, not approximate embeddings. Hallucination
risk is bounded to the synthesis stage and is zero at the knowledge-retrieval
stage.

An MCP server exposes the full federated query surface---including corpus-scoped
and person-scoped operations---to any MCP-compatible AI agent. The long-term
goal of this architecture is the \emph{TreeOfKnowledge\texttrademark}, also
termed the \emph{Corpus Scientia}: all human knowledge, all domains, one
interface, one query, coupled with the largest available language model for
grounded synthesis.
\end{abstract}

%-------------------------------------------------------------------
\section{Introduction}
%-------------------------------------------------------------------

Large software systems accumulate knowledge in incompatible forms. Source code
encodes behavior and architecture in a machine-readable but semantically dense
notation. Documentation encodes intent, rationale, and usage patterns in prose.
Domain corpora---pathway databases, ontologies, financial models, legal codes,
scripture---encode facts in formats specific to each field. These representations
are maintained separately, queried separately, and surfaced separately; an
analyst or an AI agent that needs to understand a system must traverse all of
them and reconcile the results manually.

The standard answer is retrieval-augmented generation (RAG): embed all documents
in a vector index and retrieve the closest matches to a query. This approach
treats source code, scientific data, and legal text as undifferentiated text,
discarding the formal structure that makes those artifacts tractable. A Python
call graph cannot be derived from a vector search; neither can a metabolic
pathway traversal, a statutory cross-reference chain, or a protein bond
topology. Vector retrieval approximates relevance; it does not expose structure.

An alternative, exemplified by GraphRAG~\cite{edge2024graphrag}, is to use a
language model to extract a knowledge graph from unstructured text. This
recovers some structure, but the structure is inferred rather than derived:
entity extraction is probabilistic, edges may be hallucinated, and the resulting
graph cannot be independently verified against any formal source.

KGRAG takes a different approach, grounded in a fundamental observation about
knowledge domains.

\subsection{The Ontology Principle}

The power of structural knowledge-graph retrieval scales directly with how
well-defined the ontology of the source domain is. A \emph{well-defined
ontology} is one where the entities, their relationships, and the rules
governing those relationships are formally specified---not inferred, not
approximated, but written down and enforced by construction.

Programming languages represent the ideal case. Python's grammar is a complete
formal specification; every function, class, import, and call site is expressed
in that grammar; and a deterministic parser can extract every structural
relationship without any inference. The derived call graph is not an
approximation---it is a theorem about the source code. The same property holds,
to varying degrees, across a surprising breadth of domains:

\begin{center}
\begin{tabular}{lll}
\toprule
Domain & Formal structure & KG kind \\
\midrule
Python source code & Abstract syntax tree & \texttt{code} \\
Markdown / RST documentation & Document parse tree, hyperlinks & \texttt{doc} \\
Biochemical pathways & Reaction schemas (KEGG, BioCyc) & \texttt{meta} \\
Protein structures & PDB file format (ATOM, SEQRES) & \texttt{pdbfile} \\
Disulfide bond networks & Bond topology from PDB coordinates & \texttt{disulfide} \\
Books of law & Title → Chapter → Section → Clause & \texttt{legal} \\
Diary / journal entries & Document structure + temporal order & \texttt{diary} \\
Scripture / verse & Book → Chapter → Verse, cross-refs & \texttt{verse} \\
Episodic memory & Temporal event graphs & \texttt{memory} \\
Personal knowledge & Biographical and relational graphs & \texttt{person} \\
\bottomrule
\end{tabular}
\end{center}

In every case, the formal structure of the source \emph{is} the ontology, and a
knowledge graph derived from it carries the same correctness guarantees as the
source itself. This is the foundational principle of KGRAG: build knowledge
graphs from formal structure, and the graphs are correct by construction.

\subsection{Contributions}

The contribution of this paper is not a new graph construction algorithm but a
new level of abstraction above existing single-domain tools: a uniform
federation layer, termed a \emph{knowledge compiler}, that makes heterogeneous
structural knowledge graphs queryable as a single unified corpus. Specifically:

\begin{enumerate}
  \item A five-method adapter protocol (\texttt{KGAdapter}) that decouples the
    federation layer from all backend-specific logic, enabling new knowledge
    domains to be added by implementing five methods with no changes to the
    orchestrator, registry, CLI, or MCP server.
  \item A corpus abstraction grouping multiple KG instances for scoped federated
    queries, and a person corpus abstraction extending this with personal metadata.
  \item A complete MCP server exposing registry, corpus, and person tools to
    AI agents.
  \item A demonstration that structural KG retrieval followed by language model
    synthesis eliminates hallucination at the knowledge layer, bounding risk to
    the synthesis stage alone.
  \item A long-term architectural roadmap toward the \emph{TreeOfKnowledge\texttrademark}
    (\emph{Corpus Scientia}): all human knowledge, all domains, one federated interface.
\end{enumerate}

%-------------------------------------------------------------------
\section{Design Principles}
%-------------------------------------------------------------------

Six principles govern the design of KGRAG and its backend modules.

\begin{enumerate}

\item \textbf{Derived structure is authoritative.}
  Knowledge graphs are extracted from formal sources---ASTs, document parse
  trees, ontological schemas---by deterministic programs. The graph is not a
  model of the source; it \emph{is} the source, represented relationally.
  Embeddings are derived from it and are disposable.

\item \textbf{Semantics accelerate; structure decides.}
  Vector search locates entry points. Graph traversal determines what is
  returned. The two phases are explicit, separable, and independently tunable.
  A change to the embedding model affects which seeds are selected; a change to
  the hop count or edge filter affects traversal depth. Neither affects the
  structural layer.

\item \textbf{Every result is traceable.}
  Every node carries a stable identifier encoding its origin: file path and
  qualified name for code nodes, document path and section offset for
  documentation nodes. Every snippet is extracted from a verified location.
  Nothing enters a result without a source address.

\item \textbf{Determinism over approximation.}
  Given identical inputs, the system produces identical outputs. Ranking is
  computed by a deterministic composite key. There are no stochastic components
  in the structural or traversal layers.

\item \textbf{Generality through protocol.}
  The federation layer depends on an abstract adapter interface, not on any
  specific backend. Adding a new knowledge domain requires implementing five
  methods; the orchestrator, registry, CLI, and MCP server require no changes.

\item \textbf{Independence from language models.}
  The build pipeline, query engine, graph traversal, ranking, snippet
  extraction, and analysis tools all run locally without any language model
  call. A downstream language model may consume the output, but it plays no
  role in producing it.

\end{enumerate}

%-------------------------------------------------------------------
\section{Architecture Overview}
%-------------------------------------------------------------------

The system is organized into five abstraction layers, each with a single
responsibility and depending only on layers below it. Figure~\ref{fig:layers}
shows their arrangement.

\begin{figure}[h]
\centering
\begin{tabular}{|c|}
\hline
\textbf{Layer~5: Federation} \\
\texttt{KGRAG} orchestrator, \texttt{KGRegistry}, \texttt{CorpusRegistry},
\texttt{PersonCorpusRegistry}, MCP server \\
\hline
\textbf{Layer~4: Adapter Protocol} \\
\texttt{KGAdapter}: \texttt{is\_available}, \texttt{query}, \texttt{pack},
\texttt{stats}, \texttt{analyze} \\
\hline
\textbf{Layer~3: Semantic Indexing} \\
LanceDB vector store, sentence-transformer embeddings \\
\hline
\textbf{Layer~2: Graph Storage} \\
SQLite, typed nodes and edges, BFS traversal, provenance \\
\hline
\textbf{Layer~1: Source Parsing} \\
AST (code), document parse tree (docs), ontological schema (domain) \\
\hline
\end{tabular}
\caption{KGRAG abstraction layers. Each layer depends only on those below it.
The adapter protocol at Layer~4 is the boundary between domain-specific and
domain-agnostic components.}
\label{fig:layers}
\end{figure}

Layers~1--3 are implemented independently by each knowledge graph module and
are hidden behind the Layer~4 boundary. Layer~5 operates exclusively through
the \texttt{KGAdapter} interface and has no knowledge of any module's internal
representation.

%-------------------------------------------------------------------
\section{Layer 1: Source Parsing}
%-------------------------------------------------------------------

Each knowledge graph module derives its graph from the formal structure of its
source domain. Three parsers are currently implemented; stub adapters provide
the protocol boundary for seven additional domains under development.

\subsection{Python Codebase Parsing (CodeKG)}

\texttt{CodeGraph.extract()} performs two deterministic AST passes over all
\texttt{.py} files in a repository. Pass~1 extracts definitions and emits
structural edges: \texttt{CONTAINS} (module to class, class to method),
\texttt{IMPORTS} (module-level import relationships), and \texttt{INHERITS}
(class to base class). Pass~2 extracts the call graph, emitting \texttt{CALLS}
edges annotated with call-site evidence: the line number and expression text of
each call site, extracted directly from the AST node.

A data-flow visitor (\texttt{CodeKGVisitor}) extends \texttt{ast.NodeVisitor}
to emit three additional edge types per function: \texttt{READS} (variable
consumption), \texttt{WRITES} (variable assignment), and \texttt{ATTR\_ACCESS}
(attribute access on a named object). These capture intra-function data movement
without requiring execution or type inference.

A post-build step, \texttt{resolve\_symbols()}, name-matches every unresolved
call target (represented as a \texttt{sym:} stub node) against all first-party
definitions and writes \texttt{RESOLVES\_TO} edges. This enables cross-module
fan-in queries: given a function, which functions in other modules call it?

All nine edge relations are derived from syntax. No relation is inferred.

\subsection{Document Corpus Parsing (DocKG)}

DocKG parses Markdown corpora into a heterogeneous graph. Six node kinds are
extracted: \texttt{document}, \texttt{section}, \texttt{chunk}, \texttt{entity},
\texttt{keyword}, and \texttt{topic}. Eight edge types connect them. Structural
edges follow the document hierarchy; semantic edges are derived from
co-occurrence and overlap analysis without language model involvement.

\subsection{Domain Corpus Parsing (MetaKG and Beyond)}

MetaKG wraps a domain-specific backend---initially targeting metabolic pathway
databases---via the same \texttt{KGAdapter} interface. The parsing strategy is
domain-defined; the interface contract is uniform. A biochemical pathway graph
derived from KEGG, a legal ontology derived from regulatory filings, a protein
topology derived from PDB ATOM records, and a verse graph derived from scripture
chapter/verse structure all surface identically to the federation layer.

This uniformity is not accidental. It is the central claim of the ontology
principle: \emph{where formal structure exists, structural parsing produces
graphs with the same correctness guarantees as the source, and the federation
layer need not know which domain it is querying.}

%-------------------------------------------------------------------
\section{Layer 2: Graph Storage}
%-------------------------------------------------------------------

All modules persist their graphs to SQLite in a two-table schema: a
\texttt{nodes} table and an \texttt{edges} table. Node records carry a stable
string identifier, a kind, a name, and a JSON metadata blob. Edge records carry
source node ID, relation name, destination node ID, and optional JSON evidence.

Node identifiers are deterministically constructed from kind, source path, and
qualified name. For example: \texttt{fn:src/kg\_rag/orchestrator.py:KGRAG.query}.
This means a node ID is stable across rebuilds as long as the source does not
move, and it is human-readable and directly linkable to the source.

The key traversal primitive is \texttt{expand(seed\_ids, hop, rels)}: a bounded
breadth-first search from a set of seed nodes, following a specified set of edge
relations up to a configurable hop limit. Each reached node is annotated with
\texttt{ProvMeta}: its minimum hop distance from any seed and the originating
seed node ID. Provenance travels with every result.

%-------------------------------------------------------------------
\section{Layer 3: Semantic Indexing}
%-------------------------------------------------------------------

\texttt{SemanticIndex} reads nodes from a \texttt{GraphStore}, constructs a
canonical text document for each node (name plus available descriptive text),
embeds them using a sentence-transformer model, and stores the vectors in
LanceDB. The index is derived from the SQLite graph and is fully disposable.
Its only function is to rank nodes by semantic similarity to a query string,
producing a short list of seed node IDs for the subsequent structural traversal.

The separation is intentional and load-bearing. If the embedding model is
replaced, the new index will select different seeds; the traversal from those
seeds will still follow the same structural edges and produce structurally
correct results. The accuracy of the structural layer does not depend on the
quality of the embedding model.

%-------------------------------------------------------------------
\section{Layer 4: The Adapter Protocol}
%-------------------------------------------------------------------

\texttt{KGAdapter} is an abstract base class with five abstract methods. Every
knowledge graph module exposes itself to the federation layer through this
interface and through no other.

\begin{lstlisting}[language=Python]
class KGAdapter(ABC):
    def __init__(self, entry: KGEntry) -> None: ...

    @abstractmethod
    def is_available(self) -> bool: ...

    @abstractmethod
    def query(self, q: str, k: int = 8) -> list[CrossHit]: ...

    @abstractmethod
    def pack(self, q: str, k: int = 8,
             context: int = 5) -> list[CrossSnippet]: ...

    @abstractmethod
    def stats(self) -> dict[str, Any]: ...

    @abstractmethod
    def analyze(self) -> str: ...
\end{lstlisting}

A \texttt{StubKGAdapter} base class provides a graceful default implementation
for domains whose backing library has not yet been released. Stub adapters
return empty results from \texttt{query} and \texttt{pack}, report
\texttt{status: unavailable} from \texttt{stats}, and produce a structured
unavailability notice from \texttt{analyze}. This allows new domains to be
registered, grouped into corpora, and included in federated queries before
their compilers are complete; they participate in the structure of the
knowledge graph but contribute no data until ready.

The adapter protocol defines the boundary between the universal and the
specific. Everything above Layer~4 is domain-agnostic.

%-------------------------------------------------------------------
\section{Layer 5: Federation}
%-------------------------------------------------------------------

\subsection{Registry}

\texttt{KGRegistry} maintains a persistent SQLite table of \texttt{KGEntry}
records. Each entry records the KG instance name, its kind, the repository
root, paths to its SQLite and LanceDB databases, the module version, and a
JSON metadata blob. Registration is performed once per project; subsequent
queries consult the registry to locate all known KG instances.

\subsection{Corpus and Person Corpus Registries}

Two higher-level registries group KG instances for scoped federated queries.

\texttt{CorpusRegistry} maintains a \texttt{corpora} table in the same SQLite
file. A \texttt{CorpusEntry} is a named list of \texttt{KGEntry} UUIDs with an
optional description and tags. Corpus operations (\texttt{query\_corpus},
\texttt{pack\_corpus}, \texttt{stats\_corpus}) resolve the UUID list to
\texttt{KGEntry} objects and dispatch to the same adapter machinery as
global queries.

\texttt{PersonCorpusRegistry} extends this with personal metadata: birth year,
birth date, address, email, phone, and free-form notes. A
\texttt{PersonCorpusEntry} is the natural grouping for all KGs relevant to an
individual---diary entries, memory graphs, documents, code, verse, correspondence.
Person queries fan out over all of the person's KGs:

\begin{lstlisting}
kgrag corpus person query "Eric Suchanek" \
    "disulfide bond research 2019"
\end{lstlisting}

\subsection{Orchestrator}

\texttt{KGRAG} is the cross-KG orchestrator. On each query, it:

\begin{enumerate}
  \item Retrieves all entries from the registry, optionally filtered by kind.
  \item For each entry, instantiates (or retrieves from cache) the corresponding
    \texttt{KGAdapter} and verifies availability.
  \item Dispatches the query string to each available adapter in sequence,
    collecting \texttt{CrossHit} or \texttt{CrossSnippet} lists.
  \item Merges all results into a single list and sorts by score descending.
  \item Returns a typed result object with the merged list and per-KG metadata.
\end{enumerate}

The dispatch loop is deliberately simple. Individual adapter calls are
independent; a failure in one does not abort the others. Stub adapters (for
domains whose libraries are not yet available) return empty lists and are
silently skipped.

\subsection{Result Types}

\texttt{CrossQueryResult} carries the merged hit list, the per-KG breakdown,
the total hit count, and the number of KGs that were actually queried.

\texttt{CrossSnippetPack} carries the merged snippet list and an approximate
token count for budget-aware context assembly. \texttt{CrossSnippetPack.render()}
produces a single LLM-ready string with section headers identifying each
snippet's source domain, KG name, file path, and line span.

%-------------------------------------------------------------------
\section{Query Execution}
%-------------------------------------------------------------------

A federated query executes as follows at each participating adapter:

\begin{enumerate}
  \item \textbf{Semantic seeding.} The query string is embedded and used to
    retrieve $k$ seed nodes from LanceDB.
  \item \textbf{Structural expansion.} \texttt{GraphStore.expand()} performs
    BFS from the seed IDs, following a specified set of edge relations up to
    a configurable hop limit. Each reached node is annotated with
    \texttt{ProvMeta}.
  \item \textbf{Ranking.} Nodes are ranked by a deterministic composite key:
    hop distance (ascending), seed embedding distance (ascending), node kind
    priority, node ID as tiebreaker.
  \item \textbf{Deduplication.} For snippet queries, nodes are deduplicated by
    file and source span.
  \item \textbf{Snippet extraction.} Source text is extracted at each retained
    node's verified path and line span, with a configurable context window.
\end{enumerate}

After each adapter completes, the orchestrator merges results and performs a
final global sort by score. End-to-end latency for a federated query over a
codebase of $\sim$100\,k lines is typically under one second on commodity hardware.

%-------------------------------------------------------------------
\section{Grounded Synthesis and the Hallucination Bound}
%-------------------------------------------------------------------

When the output of a KGRAG federated query is passed to a large language model
for synthesis, the combination constitutes what we term \emph{Structural
KG-RAG}~\cite{suchanek2025codekg}: knowledge-graph-augmented generation in which
the knowledge graph was not itself produced by a language model.

Every piece of context in the synthesis prompt carries:
\begin{itemize}
  \item A stable node identifier encoding its source address
  \item A source file path and line span
  \item A deterministically computed relevance score
  \item The actual source text
\end{itemize}

The language model is presented with \emph{facts}, not \emph{embeddings}. Its
role is synthesis---connecting, explaining, summarizing---not construction of
the knowledge itself. The critical consequence is that \textbf{hallucination
risk at the knowledge-retrieval stage is zero.} The model cannot fabricate a
function call that is not in the graph, because the graph was derived from the
AST; it cannot invent a statute cross-reference that is not in the legal KG,
because the legal KG was derived from the formal text; it cannot hallucinate a
protein bond that is not in the PDB file, because the bond graph was derived
from atomic coordinates.

Hallucination risk is confined to the synthesis stage alone---where the model
may reason incorrectly over correct facts. This risk is substantially lower and
is addressed by standard techniques (chain-of-thought, self-consistency,
re-querying). The knowledge foundation is provably correct by construction.

This is the decisive advantage of structural KG-RAG over inference-based
approaches: the correctness guarantee is unconditional with respect to the
structural layer and does not depend on prompt engineering, model scale, or
retrieval quality.

%-------------------------------------------------------------------
\section{The TreeOfKnowledge\texttrademark: Corpus Scientia}
%-------------------------------------------------------------------

The architecture described in this paper is a foundation for a larger goal:
the \emph{TreeOfKnowledge\texttrademark}, also termed the \emph{Corpus Scientia}.

\begin{quote}
\textit{All human knowledge. All domains. One interface. One query.}
\end{quote}

The ontology principle establishes that any domain with well-defined formal
structure can be compiled into a knowledge graph using the KGRAG adapter
protocol. The federation layer then makes all such graphs queryable as a
single unified corpus. As domain-specific compilers (DiaryKG, LegalKG,
DisulfideKG, PDBFileKG, VerseKG, MemoryKG, PersonKG) are built and
registered, the system progressively approaches completeness across domains:

\begin{itemize}
  \item \textbf{Code corpora} across all first-party software projects
  \item \textbf{Scientific literature} structured by citation graphs and
    domain ontologies
  \item \textbf{Legal codes} structured by the hierarchical organization of
    statutes, regulations, and case law
  \item \textbf{Biological structure} structured by PDB file format and
    pathway reaction schemas
  \item \textbf{Personal knowledge} structured by diary timelines, memory
    graphs, and biographical records
  \item \textbf{Cultural and religious texts} structured by the canonical
    book/chapter/verse hierarchies of scripture and the formal apparatus of
    commentary traditions
\end{itemize}

Coupled with the largest available language model and maximum context window,
a query to the Corpus Scientia retrieves the most relevant subgraph across
\emph{all} registered knowledge, passes it to the model with full provenance,
and receives a synthesis that is grounded, traceable, and verifiably correct
at the knowledge layer. This is not retrieval-augmented generation; it is
\emph{knowledge-compiler-augmented synthesis}.

The architecture is ready. The adapter protocol requires five methods per new
domain. The orchestrator, registry, CLI, MCP server, corpus, and person
abstractions require no changes as new domains are added. The path to the
Corpus Scientia is, in engineering terms, incremental: one domain compiler
at a time, each following the same pattern, each extending the unified query
surface without modifying the federation layer.

%-------------------------------------------------------------------
\section{Snapshot System}
%-------------------------------------------------------------------

Knowledge graphs change as their source corpora evolve. The snapshot system
captures \texttt{SnapshotMetrics} at each commit, keyed by the pre-commit tree
hash produced by \texttt{git write-tree}. A \texttt{Snapshot} record carries
the tree hash, branch name, timestamp, module version, a \texttt{SnapshotMetrics}
struct (node counts, edge counts, coverage score, issue count), and
\texttt{SnapshotDelta} fields comparing against the immediately preceding and
baseline snapshots.

The snapshot history provides a longitudinal record of graph evolution: which
commits added nodes, which removed them, and how semantic coverage trended over
time.

%-------------------------------------------------------------------
\section{MCP Integration}
%-------------------------------------------------------------------

An MCP server wraps the \texttt{KGRAG} orchestrator and exposes a complete
tool surface to any MCP-compatible AI agent. Tools are organized into three
groups.

\textbf{Registry tools:}
\begin{center}
\begin{tabular}{ll}
\toprule
Tool & Function \\
\midrule
\texttt{kgrag\_stats()} & Registry summary: KG count, kinds, built status \\
\texttt{kgrag\_list([kind])} & List registered KG entries, optionally filtered \\
\texttt{kgrag\_info(name)} & Full detail for a single KG entry \\
\texttt{kgrag\_query(q, [k, kinds])} & Federated semantic query, JSON result \\
\texttt{kgrag\_pack(q, [k, kinds])} & Federated snippet pack, Markdown output \\
\bottomrule
\end{tabular}
\end{center}

\textbf{Corpus tools} (\texttt{kgrag\_corpus\_list}, \texttt{kgrag\_corpus\_info},
\texttt{kgrag\_corpus\_create}, \texttt{kgrag\_corpus\_delete},
\texttt{kgrag\_corpus\_add}, \texttt{kgrag\_corpus\_remove},
\texttt{kgrag\_corpus\_query}, \texttt{kgrag\_corpus\_pack}): create and manage
named collections of KG instances and execute corpus-scoped federated queries.

\textbf{Person tools} (\texttt{kgrag\_person\_list}, \texttt{kgrag\_person\_info},
\texttt{kgrag\_person\_create}, \texttt{kgrag\_person\_delete},
\texttt{kgrag\_person\_add}, \texttt{kgrag\_person\_remove},
\texttt{kgrag\_person\_update}, \texttt{kgrag\_person\_query},
\texttt{kgrag\_person\_pack}): create and manage person corpus entries with
personal metadata and execute person-scoped federated queries.

The MCP transport is \texttt{stdio}, making it compatible with Claude Code,
Kilo Code, GitHub Copilot, Cline, and Claude Desktop without additional
networking configuration.

%-------------------------------------------------------------------
\section{Comparison with Inference-Based Graph Construction}
%-------------------------------------------------------------------

GraphRAG~\cite{edge2024graphrag} is the most widely deployed system in the
vicinity of this work. Both KGRAG and GraphRAG augment retrieval with a
knowledge graph; the origin and guarantees of that graph differ substantially.

In GraphRAG, a language model extracts entities and relationships from text
documents. The graph is a product of language model inference; its correctness
is bounded by extraction prompt quality and model comprehension. Errors
introduced during extraction are silent: the system has no mechanism to verify
that an extracted relationship corresponds to anything in the source.

In KGRAG, knowledge graphs are derived from formal artifacts by deterministic
programs. A \texttt{CALLS} edge exists because a call expression was found in
the AST at the recorded line number. A metabolic reaction edge exists because
it appears in the source schema. A statutory cross-reference edge exists because
a citation was found in the formal text. No edge is inferred; every edge is
verified against a source address.

\begin{table}[h]
\centering
\caption{Inference-based vs.\ structure-derived knowledge graph retrieval}
\label{tab:comparison}
\begin{tabular}{lll}
\toprule
Property & GraphRAG~\cite{edge2024graphrag} & KGRAG \\
\midrule
Graph source        & LLM extraction from text      & Formal source parsing \\
Graph correctness   & Probabilistic                 & Deterministic \\
Edge evidence       & NL relationship description   & Source address + line \\
LLM required to build  & Yes                        & No \\
LLM required to query  & No                         & No \\
Domain generality   & Text corpora                  & Any formally structured domain \\
Hallucination risk (KG layer) & Present             & Zero \\
Hallucination risk (synthesis) & Present            & Present (bounded) \\
Result provenance   & Document reference            & Node ID + file + line \\
Query latency       & Seconds to minutes            & Under one second (local) \\
\bottomrule
\end{tabular}
\end{table}

The two systems are not in competition for the same problem. GraphRAG is
designed for global sensemaking over large unstructured text corpora. KGRAG
is designed for structural navigation of formally specified artifacts, with a
zero-hallucination guarantee at the knowledge layer.

%-------------------------------------------------------------------
\section{Conclusion}
%-------------------------------------------------------------------

KGRAG demonstrates that structural knowledge from heterogeneous source domains
can be federated under a single uniform interface without collapsing domain
specificity into a common text representation. The adapter protocol cleanly
separates domain-specific extraction from domain-agnostic orchestration.

The central findings of this work are:

\begin{enumerate}

\item \textbf{The Ontology Principle.} The power of structural KG retrieval
  scales with the formal completeness of the source domain's ontology.
  Programming languages are the ideal case; biochemical pathways, legal codes,
  protein structure files, and scripture hierarchies follow the same pattern.
  Where formal structure exists, the derived graph is correct by construction.

\item \textbf{The Knowledge Compiler.} KGRAG is not a retrieval system that
  happens to use graphs. It is a knowledge compiler: it transforms formally
  structured source artifacts into relational representations that preserve and
  expose the structure of the originals. The compiled knowledge graph \emph{is}
  the corpus, represented for navigation.

\item \textbf{Universal Federation.} Because all domain adapters expose the
  same five-method interface, all compiled knowledge is queryable through one
  interface. The scientist, lawyer, software engineer, and biographer query the
  same system. One command, one ranking, full provenance.

\item \textbf{Zero Hallucination at the Knowledge Layer.} When KGRAG output is
  passed to a language model for synthesis, the model receives verified facts
  with source addresses. Hallucination risk is bounded to the synthesis stage
  alone and is zero at the knowledge-retrieval stage. This is an unconditional
  guarantee: it holds regardless of model scale, prompt engineering, or
  retrieval quality.

\item \textbf{The Path to the Corpus Scientia.} The \emph{TreeOfKnowledge\texttrademark}
  is achievable. Each new domain requires five adapter methods. The federation
  layer requires no changes. The architecture is ready; the remaining work is
  building the domain compilers.

\end{enumerate}

The result is a retrieval system that is fast, auditable, and correct by
construction at the structural layer. When coupled with the largest available
language model for synthesis, it provides grounded context whose every element
can be verified against the source. The correctness guarantee holds regardless
of the language model's accuracy, because the model receives facts rather than
constructs them. That is the ultimate power of the design.

%-------------------------------------------------------------------
\begin{thebibliography}{9}

\bibitem{edge2024graphrag}
Darren Edge, Ha Trinh, Newman Cheng, Joshua Bradley, Alex Chao, Apurva Mody,
Steven Truitt, and Jonathan Larson.
\newblock From local to global: A graph {RAG} approach to query-focused
summarization.
\newblock \emph{arXiv preprint arXiv:2404.16130}, 2024.
\newblock \url{https://arxiv.org/abs/2404.16130}

\bibitem{suchanek2025codekg}
Eric G. Suchanek.
\newblock {CodeKG}: Deterministic code navigation via knowledge graphs,
semantic indexing, and grounded context packing.
\newblock Technical report, Flux-Frontiers, 2025.
\newblock \url{https://github.com/Flux-Frontiers/code\_kg}

\bibitem{reimers2019sbert}
Nils Reimers and Iryna Gurevych.
\newblock Sentence-{BERT}: Sentence embeddings using {Siamese} {BERT}-networks.
\newblock In \emph{Proceedings of EMNLP 2019}, 2019.
\newblock \url{https://arxiv.org/abs/1908.10084}

\end{thebibliography}

\end{document}
